---
title: "A2"
author: "Zhiheng Zhou"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# CH 5

## 5.6 (p285)

```{r}
install.packages("astsa")
install.packages("fGarch")
install.packages("vars")
```

```{r}
library(astsa)
library(fGarch)
# define the growth rate of weekly oil price
growth_rate_oil <- diff(log(oil))
# fit the squares of the residuals from AR(1)
mod.ar <- arima(growth_rate_oil, order = c(1,0,0))
acf2(growth_rate_oil, 40)
acf2((mod.ar$residuals)^2, 40)
acf2(growth_rate_oil^2, 40)
```

The ACF and PACF of the data of weekly oil price growth rate look like ARMA(1,1) since both ACF and PACF decay slowly. The ACF and PACF of the squares of residuals also show ARMA pattern. Therefore, the growth rate of weekly oil prices exhibits GARCH behavior. The ACF and PACF of the squared series also show ARMA(1,1) pattern. This suggests a GARCH(1,1) model.


```{r}
# fit a GARCH(1,1) model
mod.garch <- garchFit(~arma(1, 1) + garch(1, 1), growth_rate_oil, include.mean=F)
round(mod.garch@fit$matcoef,3)
```

Noticed that the p-values all indicate statistical significance. Therefore, we fit the squares of the growth rate of weekly oil prices to an ARMA(1,1)-GARCH(1,1) model.

# 5.7 (p285)


We choose Germany DAX series for this question.

The general steps to build a model for time series process is:

- Plotting the data
- Transforming the data if necessary
- Identifying the dependence orders
- Parameter estimation
- Diagnostics
- Model selection


## Plot the data

```{r}
# plot the data
plot(EuStockMarkets[,"DAX"])
```
We notice that the data of Germany DAX indices have upwarding trends and larger variances at the tails. We do not see clear seasonality in this case. 

## Transform the data

```{r}
# Transform the data
data.dax <- diff(log(EuStockMarkets[,"DAX"]))
plot(data.dax)
```
We take first difference to control the trend and take logarithm to control the variances. We see now the data looks like white noise except for some outliers before 1992. We now use the differenced data for following steps. Note that the transformed data is as an approximation of the growth rate growth rate the weekly oil price, which is required by the question. 

## Identify the dependence orders

```{r}
# plot acf and pacf of data
acf2(data.dax, 100)
```

We see no significant autocorrelation in the series. Since the data behaves like white noise, it can be conditionally heteroskedastic. Thus, we then plot the ACF and PACF for the squares of residuals to determine a GARCH model.

```{r}
acf2(data.dax^2, 40)
```

We see that the PACF cuts off after one lag. Therefore, we fit the DAX growth rate data to a GARCH(1,0) model. By looking at the ACF and PACF, we may also consider fitting a GARCH(1,1) model. We see that both models show small p-values on the estimates. The estimates are statistically significant. 

## Parameter estimation

```{r}
#fit GARCH model
dax.garch.1 <- garchFit(~ garch(1, 1), data.dax, include.mean=F)
round(dax.garch.1@fit$matcoef,3)
dax.garch.2 <- garchFit(~ garch(1, 0), data.dax, include.mean=F)
round(dax.garch.2@fit$matcoef,3)
```

Since the estimates are all significant, fitting a GARCH(1,1) model gives us $\alpha_0 = 0, \alpha_1 = 0.068, \beta_1 = 0.889$; and fitting a GARCH(1,0) model gives us $\alpha_0 = 0, \alpha_1 = 0.097$. Notice that fitting a GARCH(1,0) does not include the GARCH term, and thus, there is no $\beta$. In this case, the variance at time t does not depend on the previous variance. 

## Diagnosis

```{r}
# model summary
summary(dax.garch.1)
summary(dax.garch.2)
```

## Model selection

According to the model summary, we may choose GARCH(1,1) model because it has lower AIC and BIC than the other. It also has higher log likelihood. All of its three estimates are statistically significant. The final model is GARCH(1,1). The Ljung-Box Test shows that there is no significant autocorrelation in the residuals. However, neither of the two models shows a normality distribution in residuals according to the results of Jarque-Bera Test and Shapiro-Wilk Test. It means that neither of them is the best model for this time series. We need to try more models.


# 5.12(p287)


```{r}
library(vars)
x <- cbind(log(econ5[,"unemp"]), log(econ5[,"gnp"]), log(econ5[,"consum"]))
VARselect(x, lag.max = 10, type = "both")
```

We first decide the order of the model. The BIC selects order p = 2, and the AIC, FPE and Hannan-Quinn selects order p = 3.

We can select p = 2

```{r}
fit <- VAR(x, p = 2, type = "both")
summary(fit$varresult$log.econ5....unemp...)
summary(fit$varresult$log.econ5....gnp...)
summary(fit$varresult$log.econ5....consum...)
```

We can summarize that, $\hat{\alpha} = (3.84, -0.24, -0.08)', \hat{\beta} = (0.004, 0.0004, 0)$, where $\hat{\alpha}$ is the constant and $\hat{\beta}$ is the trend.

```{r}
acf(resid(fit), 50)
```

The ACFs of the individual residual series are along the diagonal. The CCFs are displayed off diagonals. We notice that unemployment/GNP and consumption/GNP are above the significance level. It means that the model does not capture the concurrent effects of unemployment and consumption on GNP. We need to try more models.